Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-04-29 manuscript/manuscript.tex
updated histograms in Sec. II

The optimal statistic is constructed from the autocovariance and cross-covariance matrices of the residuals 
${\bf{P}}_a$ and ${\bf{S}}_{ab}$,
\begin{eqnarray}
	{\bf{P}}_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	{\bf{S}}_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,.
\end{eqnarray}
where $a,b$ index pulsars in the PTA. 
%In the frequency domain, these can be written as
%\begin{eqnarray}
%	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
%	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
%\end{eqnarray}
%where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
%two observations, 
%$\mathbf{R}$ is the timing model fitting matrix, 
%$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
%$P_\mathrm{gw}(f)$ is the GW power spectrum, 
%and $P_I(f)$ is the power spectrum for the $I$th pulsar. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) } \,,
\end{equation}
where $\tilde{{\bf{S}}}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{{\bf{S}}}_{ab} = {\bf{S}}_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
For a weak GW signal, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{-1/2} \,,
\end{equation}
and the signal-to-noise ratio (SNR) is
\begin{equation}
	\rho = \frac{\hat{A}^2}{\sigma_0} = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2}} \,.
\end{equation}
The expectation value of the SNR, averaged over all realizations of the stochastic background, is
\begin{equation}
	\left\langle \rho \right\rangle = \Agw^2 \left[ \sum_{ab} \Tr\left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2} \,.
\end{equation}

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).
\begin{table}[t]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs. \sv{finish filling in this table}}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a simulation with a GWB with $\Agw=5\times10^{-15}$. 
%We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%for a common red process. 
For this particular realization of the GWB, 
the fixed-noise analysis 
using the individual noise results gives 
$\hat{A}^2 = 6.6\times10^{-30}$ with $\mathrm{SNR} = 2.4$, 
and the fixed-noise analysis using the common noise results gives 
$\hat{A}^2 = 2.6\times10^{-29}$ with $\mathrm{SNR} = 6.0$. 
The noise-marginalized analysis gives $\hat{A}^2 = (2.5 \pm 0.1)\times10^{-29}$ 
with $\mathrm{SNR}=4.8\pm0.8$. 
The value of $\hat{A}^2$ from the fixed-noise analysis using the individual noise results 
is significantly lower than the injected level of the GWB, while 
the values of $\hat{A}^2$ from the fixed-noise analysis using the common noise results 
and the noise-marginalized analysis are in good agreement with each other and the injected value. 
The fixed-noise analysis using the individual noise results also gives a significantly lower SNR 
than the analyses using the common noise results. 
The mean SNR from the noise-marginalized analysis is smaller 
than the SNR found with the fixed-noise analysis using the common noise results, 
but the two are consistent at $1.6\sigma$.
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic for a simulated dataset containing an injected GWB. 
			We compare the values found using a fixed-noise analysis with noise parameters taken from 
			individual and common noise analyses (dashed blue lines and solid orange lines, respectively) 
			to the values found by marginalizing over 
			the individual pulsars' red noise parameters (solid green lines). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the GWB, the fixed-noise analysis 
			using the individual noise results underestimates the strength of the GWB, 
			while the fixed-noise analysis using the common noise results 
			and the noise-marginalized analysis more accurately find $\hat{A}^2$.}
	\label{fig:os_dataset_sample}
\end{figure}
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
%	\caption{Optimal statistic and SNR for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters (solid orange line). 
%			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%			for a common red process (solid green line). 
%			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			For this particular realization of the stochastic background, the fixed-noise analysis 
%			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
%			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
%			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
%	\label{fig:os_dataset_sample}
%\end{figure}

@@

The fixed-noise and noise-marginalized analyses give the same results 
for an injected GWB with $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
Figure~\ref{fig:os_datasetstats2} shows the mean optimal statistic and mean SNR 
for 100 realizations of a GWB with $\Agw = 10^{-15}$. 
The fixed-noise analysis using the individual noise values 
gives a mean value of $\hat{A}^2 = 2.9\times10^{-31}$ and a mean SNR of 1.1. 
The fixed-noise analysis using the common noise values 
gives a mean value of $\hat{A}^2 = 8.4\times10^{-31}$ and a mean SNR of 2.1, 
which the noise-marginalized analysis gives a mean value of 
$\hat{A}^2 = 1.1\times10^{-30}$ and a mean SNR of 2.4.
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_A1e-15.pdf}
	\caption{Optimal statistic and SNR for 100 simulated datasets 
			containing an injected GWB with $\hat{A}^2 = 10^{-30}$ (dashed vertical line). 
			We show the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			to the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
			that are closer to the injected value.}
	\label{fig:os_datasetstats2}
\end{figure}

We also compare these three methods 
based on how far the computed values of the optimal statistic are from the injected value of of the GWB 
compared to the uncertainty in the optimal statistic,
\begin{equation}
	s = \frac{\hat{A}^2 - \Agw^2}{\sigma} = \frac{\hat{A}^2 - \Agw^2}{\hat{A}^2/\rho} \,.
\end{equation}
Figure~\ref{fig:os_compare} shows the distribution of $(\hat{A}^2-\Agw^2)/\sigma$

>>>>

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
${\bf{P}}_a$ and ${\bf{S}}_{ab}$,
\begin{eqnarray}
	{\bf{P}}_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	{\bf{S}}_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,.
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
%In the frequency domain, these can be written as
%\begin{eqnarray}
%	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
%	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
%\end{eqnarray}
%where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
%two observations, 
%$\mathbf{R}$ is the timing model fitting matrix, 
%$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
%$P_\mathrm{gw}(f)$ is the GW power spectrum, 
%and $P_I(f)$ is the power spectrum for the $I$th pulsar. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) } \,,
\end{equation}
where $\tilde{{\bf{S}}}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{{\bf{S}}}_{ab} = {\bf{S}}_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\hat{A}^2}{\sigma_0} = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2}} \,.
\end{equation}
The expectation value of the SNR, averaged over all realizations of the stochastic background, is
\begin{equation}
	\left\langle \rho \right\rangle = \Agw^2 \left[ \sum_{ab} \Tr\left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2} \,.
\end{equation}

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).
\begin{table}[tb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs. \sv{finish filling in this table}}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a simulation with a GWB with $\Agw=5\times10^{-15}$. 
%We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%for a common red process. 
For this particular realization of the GWB, 
the fixed-noise analysis 
using the individual noise results gives 
$\hat{A}^2 = 6.6\times10^{-30}$ with $\mathrm{SNR} = 2.4$, 
and the fixed-noise analysis using the common noise results gives 
$\hat{A}^2 = 2.6\times10^{-29}$ with $\mathrm{SNR} = 6.0$. 
The noise-marginalized analysis gives $\hat{A}^2 = (2.5 \pm 0.1)\times10^{-29}$ 
with $\mathrm{SNR}=4.8\pm0.8$. 
The value of $\hat{A}^2$ from the fixed-noise analysis using the individual noise results 
is significantly lower than the injected level of the GWB, while 
the values of $\hat{A}^2$ from the fixed-noise analysis using the common noise results 
and the noise-marginalized analysis are in good agreement with each other and the injected value. 
The fixed-noise analysis using the individual noise results also gives a significantly lower SNR 
than the analyses using the common noise results. 
The mean SNR from the noise-marginalized analysis is smaller 
than the SNR found with the fixed-noise analysis using the common noise results, 
but the two are consistent at $1.6\sigma$.
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic for a simulated PTA dataset containing an injected GWB. 
			We compare the values found using a fixed-noise analysis 
			with noise parameters taken from 
			individual and common noise analyses 
			(dashed blue lines and solid orange lines, respectively) 
			to the values found by marginalizing over 
			the individual pulsars' red noise parameters (green histograms). 
			The dotted vertical line indicates the injected value, $\Agw^2 = 2.5 \times 10^{-29}$. 
			The fixed-noise analysis 
			using the individual noise results underestimates $\Agw$, 
			while the fixed-noise analysis using the common noise results 
			and the noise-marginalized analysis more accurately recover $\Agw$.}
	\label{fig:os_dataset_sample}
\end{figure}
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
%	\caption{Optimal statistic and SNR for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters (solid orange line). 
%			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%			for a common red process (solid green line). 
%			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			For this particular realization of the stochastic background, the fixed-noise analysis 
%			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
%			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
%			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
%	\label{fig:os_dataset_sample}
%\end{figure}

@@

The fixed-noise and noise-marginalized analyses give the same results 
for an injected GWB with $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
%Figure~\ref{fig:os_datasetstats2} shows the mean optimal statistic and mean SNR 
%for 100 realizations of a GWB with $\Agw = 10^{-15}$. 
%The fixed-noise analysis using the individual noise values 
%gives a mean value of $\hat{A}^2 = 2.9\times10^{-31}$ and a mean SNR of 1.1. 
%The fixed-noise analysis using the common noise values 
%gives a mean value of $\hat{A}^2 = 8.4\times10^{-31}$ and a mean SNR of 2.1, 
%which the noise-marginalized analysis gives a mean value of 
%$\hat{A}^2 = 1.1\times10^{-30}$ and a mean SNR of 2.4.
%\begin{figure}[t]
%	\includegraphics[width=0.95\columnwidth]{plots/optstat_A1e-15.pdf}
%	\caption{Optimal statistic and SNR for 100 simulated datasets 
%			containing an injected GWB with $\hat{A}^2 = 10^{-30}$ (dashed vertical line). 
%			We show the values found by fixing the red noise 
%			based on the results from individual and common noise analyses 
%			(blue and orange lines, respectively) 
%			to the mean optimal statistic and mean SNR found by marginalizing over 
%			the red noise parameters using the results of a common noise analysis (green lines). 
%			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
%			that are significantly lower than the injected value, 
%			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
%			that are closer to the injected value.}
%	\label{fig:os_datasetstats2}
%\end{figure}
We also compare these three methods 
based on how far the computed values of the optimal statistic are from the injected value of of the GWB 
compared to the uncertainty in the optimal statistic,
\begin{equation}
	s = \frac{\hat{A}^2 - \Agw^2}{\sigma} = \frac{\hat{A}^2 - \Agw^2}{\hat{A}^2/\rho} \,.
\end{equation}
Figure~\ref{fig:os_compare} shows the distribution of $(\hat{A}^2-\Agw^2)/\sigma$
\begin{figure*}[t]
	\includegraphics[width=0.95\columnwidth]{plots/pp_plot.pdf}
	\caption{\sv{write caption}}
	\label{fig:os_compare}
\end{figure*}
